Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-04-02 paper.tex
more physical arguments in intro

\section{Thermodynamic requirement for the vertical shear instability}\label{vsi_require}    
In this paper we will find that in a stably stratified disk, the VSI
requires a \emph{very} short thermal timescale $t_c$, much less than the
disk dynamical timescale,
\begin{align}
  t _c \ll \Omega^{-1},  
\end{align}
consistent with numerical experiments \citep{nelson13}. In subsequent
sections, we quantify this requirement more 
precisely. For this preliminary discussion we shall 
consider a vertically isothermal disk with aspect-ratio $\epsilon$,
dimensionless radial temperature gradient $q$ and adiabatic index
$\gamma$.

@@

However, the stabilizing effect of buoyancy can be mitigated for
non-adiabatic disturbances. Consider a fluid element displaced upwards, away from
the disk midplane. In a stably stratified adiabatic disk, it would
sink because of buoyancy. However, if thermal timescales are short, in
that temperature flucations decay rapidly (i.e. non-adiabatic), then
the fluid element would attain its surrounding density and 
%(assuming an ideal gas and it reaches pressure 
%equilibrium with its surroundings). 
thus buoyancy is rendered ineffective. Qualitatively, a short thermal
timescale $t_c$ reduces the effective (stabilizing) buoyancy force.

How short must $t_c$ be for vertical shear to prevail? We see
from above that the destabilizing effect of vertical shear is rather
weak because $|\p_z\Omega|\propto \epsilon \ll 1$. 
Then even a small amount of buoyancy can significantly stabilize
the VSI. It follows that $t_c$ must be \emph{extremely} short. The
central result of this work is the requirement 
\begin{align}\label{prelim_bcrit}
  (\gamma - 1)t_c < \epsilon |q| \Omega^{-1}
\end{align}   
for the VSI in a vertically isothermal, thin disk. Physically, the  
left-hand side represents an effective buoyancy, while the right-hand
side represents the (weakly)  destabilizing effect of vertical
shear. \emph{In a thin disk the thermal timescale must be shorter than
  the dynamical timescale by $O(\epsilon)\ll 1$ for the VSI to operate.}

@@

The simplicity of Eq. \ref{prelim_bcrit} means it can be 
readily applied to evaluate where in a disk the VSI is likely to
operate. For example, we show in Fig. \ref{bcrit_mmsn2d} an estimate
of the thermal timescales in a typical protoplanetary disk model (see 
below). Here, $H$ is the disk scale height and $k_x$ is a radial
wavenumber of the disturbance being considered, thus larger $k_x$
correspond to smaller perturbations. The region outside the red line
satisfies Eq. \ref{prelim_bcrit}.

The derivation of Eq. \ref{prelim_bcrit} in fact
assumes $t_c$ is vertically uniform. While this is not generally true
in the inner disk, we find Eq. \ref{prelim_bcrit} still provides a
useful guide. We see that the VSI is likely to  operate in the outer 
disk at tens of AU, but that smaller scale disturbances may operate at
smaller radii.

>>>>

\section{Thermodynamic requirement for the vertical shear instability}\label{vsi_require}    
In this paper we will find that in a stably stratified disk, the VSI
requires a \emph{very} short thermal timescale $t_c$, much less than the
disk dynamical timescale,
\begin{align}
  t _c \ll \Omega^{-1},  
\end{align}
consistent with numerical experiments \citep{nelson13}. For this
preliminary discussion we shall consider a vertically isothermal disk
with aspect-ratio $\epsilon$, 
dimensionless radial temperature gradient $q$ and adiabatic index
$\gamma$.

@@

However, the stabilizing effect of buoyancy can be mitigated for
non-adiabatic disturbances. Consider a fluid element displaced upwards, away from
the disk midplane. In a stably stratified adiabatic disk, it would
sink because of buoyancy. However, if thermal timescales are short, in
that temperature flucations decay rapidly (i.e. isothermal perturbations), then 
the fluid element would attain its surrounding density and 
%(assuming an ideal gas and it reaches pressure 
%equilibrium with its surroundings). 
thus buoyancy is rendered ineffective. Qualitatively, a short thermal
timescale $t_c$ reduces the effective (stabilizing) buoyancy force.

How short must $t_c$ be for vertical shear to prevail? For a rough
comparison of vertical shear to buoyancy for non-adiabatic
perturbations, we apply Eq. 47 of \cite{urpin03}, which is based on
\cite{townsend58},   
\begin{align}\label{bcrit_est}
  N_z^2 t_c \lesssim \left|r\frac{\p\Omega}{\p z}\right|. 
\end{align}
The left-hand side represents an effective buoyancy, which is
stabilizing; while the right-hand side represents vertical shear,
which is destabilizing. In a thin disk, the destabilizing effect of
vertical shear is rather weak because $|\p_z\Omega|\propto \epsilon
\ll 1$. Then even a small amount of buoyancy can significantly
stabilize the VSI. It follows that $t_c$ must be quite short to permit
instability.

Note that in astrophysical disks both $N_z^2$ and $\p_z\Omega$ varies
with height. Eq. \ref{bcrit_est} then suggest that with 
increasing height, buoyancy would eventually stabilize the VSI (since
$N_z^2\propto z^2$ and $\p_z\Omega\propto z$). 
%This means that 
%disturbances localized away from the disk midplane are easily stabilized by
%buoyancy. 
This argument assumes vertically-localized disturbances. It is,
however, not clear how to apply Eq. \ref{bcrit_est} to vertically
global disturbances, as considered in this work.

%it would imply you can always have VSI sufficiently close to the
%midplane

Our central result is the requirement 
\begin{align}\label{prelim_bcrit}
  (\gamma - 1)t_c < \epsilon |q| \Omega^{-1}
\end{align}   
for the VSI in a vertically isothermal, thin disk. This equation
reflects Eq. \ref{bcrit_est} but is simpler since there is no
$z$ dependence. We find that \emph{in a thin 
  disk the thermal timescale must be shorter than the dynamical
  timescale by $O(\epsilon)\ll 1$ for the VSI to operate.}

@@

Eq. \ref{prelim_bcrit} can be 
readily applied to evaluate where in a disk the VSI is likely to
operate. For example, we show in Fig. \ref{bcrit_mmsn2d} an estimate
of the thermal timescales in a typical protoplanetary disk model (see 
below). Here, $H$ is the disk scale height and $k_x$ is a radial
wavenumber of the disturbance being considered, thus larger $k_x$
correspond to smaller perturbations. The region outside the red line
satisfies Eq. \ref{prelim_bcrit}. The criterion Eq. \ref{prelim_bcrit} in fact
assumes $t_c$ is vertically uniform. While this is not generally true
in the inner disk, we find Eq. \ref{prelim_bcrit} still provides a
useful guide. We see that the VSI is likely to  operate in the outer 
disk at tens of AU, but that smaller scale disturbances may operate at
smaller radii.
